Optimistic Atomic Multicast in One Communication Delay

### 2011-08-02 textos/optpaxos/draft_dropbox.tex
text revised, missing grammar and typo polishing

To do the optimistic delivery, after a new message $m$ has been \rmdel{}ed at $p$, instead of proposing it, $p$ inserts it into an $optPending$ set. Then, $p$ waits until its own wallclock has a value greater than $m.ts + w(p)$. At this moment, if \emph{A2} holds, then all messages that could possibly have a timestamp lower than $m$ have already been received sent by $p$. Therefore, after $w(p)$ such messages can be \optdel{}ed in the order of their initial timestamps. Besides, they can be proposed by every process in this same order, in which they are also decided, meaning that there will be no timestamp changes.

The optimistic assumption allows for further improvement. If it holds, the timestamp of the messages will never be changed. Therefore, if barriers are being requested for each message $m$, as described in Section \ref{sec:nullondemand}, the requests may be sent to $blockers(m)$ at the same time that $m$ is \rmcast{} by its sender.

If \emph{A2} fails to hold and some message $m$ is received by a process $p$ after a message \mbox{$m' : m'.ts > m.ts$} has been already \optdel{}ed and proposed by $p$, then the optimistic delivery algorithm made a \textit{mistake}, which the application can figure out by the difference between the optimistic delivery order and the conservative one. The application should then take action to correct whatever problem this mistake might have caused. Besides, when such a mistake happens, it means that the timestamp of the message has been changed by the conservative delivery algorithm and that, maybe, a new barrier request must be sent to $blockers(m)$ when $m$'s new timestamp is defined.

@@

\subsection{Composition of the timestamp}
\label{sec:compositets}

@@

Although unlikely to happen, the worst case delivery time of \amcast{} when using Paxos with a leader for consensus and requesting barriers in the beginning of the protocol is $6\delta$. It is so because some messages may need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. However, a simple way to eliminate the possibility of having to change timestamps is by not considering messages whose delivery order does not follow the order of the initial timestamps.

Although discarding messages which do not follow the initial timestamp order conflicts with the properties we have defined for the \amcast{} primitive, we describe it here because then every message would be delivered within $3\delta$, in the worst case. The validity property would now be changed to:

\textbf{Optimistic Uniform Validity}: if a process \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then one of the correct processes that is a destination of $m$ eventually \consdel{}s $m$.

>>>>

To do the optimistic delivery, after a new message $m$ has been \rmdel{}ed at $p$, instead of proposing it, $p$ inserts it into an $optPending$ set. Then, $p$ waits until its own wallclock has a value greater than $m.ts + w(p)$. At this moment, if \emph{A2} holds, then all messages that could possibly have an initial timestamp lower than $m$ have already been received by $p$. Therefore, such messages can be \optdel{}ed in the order of their initial timestamps.

If every process $q$ of $g$ have estimated $w(q)$ correctly, all of them will propose all messages from $g$ in the same order, which is that of the initial timestamps. If this happens, there will be no timestamp changes and the optimistic delivery will be correct. Not changing timestamps allows for further improvement: if barriers are being requested for each message $m$, as described in Section \ref{sec:nullondemand}, the barrier requests may be sent to $blockers(m)$ at the same time as $m$ is \rmcast{} by its sender.

If \emph{A2} fails to hold and some message $m$ is received by some process $p$ after a message $m'$ has been already \optdel{}ed by $p$, where $m.ts < m'.ts$, then the optimistic delivery algorithm has made a \textit{mistake}. Such mistake can be figured out by the application from the difference between the optimistic delivery order and the conservative one. The application should then correct whatever problems this mistake might have caused. Besides, when such a mistake happens, it means that the timestamp of a message may have changed and that a new barrier request may have to be sent to $blockers(m)$ once the new timestamp has been defined.

@@

\subsection{Timestamp composition}
\label{sec:compositets}

\textit{\textbf{I'm not sure of how good this `optimization' is; maybe it's not worth putting on the paper...}}

Barrier requests for each message $m$ may be sent before $m$'s final timestamp is decided. However, if $m.ts$ changes, new barrier request will probably be deemed necessary. For instance, let $m$ and $m'$ be two messages sent by different processes of the same group $g$, where their initial timestamps are such that $m.ts < m'.ts$. Assume also that barrier requests have been sent at the same time as such messages were being sent. Suppose that some process $p$ from $g$, due to some asynchronous behavior of the system, received $m'$ first, proposed it and had $m'$ decided before $m$. As $m$ will be decided afterwards, its timestamp will be made greater than that of $m'$. Therefore, the barriers requested in order to deliver $m$ might not suffice anymore, requiring new, higher ones.

An approach to reduce this problem could be by using composite timestamps. Instead of relying on a single value -- apart from the message id for breaking ties --, two values could be used: one related to the instant when the message was created and one sequence number incremented only when the timestamp should be increased. The former we call $rtc$, standing for real-time clock, and the latter we call $seq$. Being $m$ any message, $m.ts = (rtc,seq)$. The second value is used only to break ties when the first one is equal between some two messages. In the case when the timestamp of $m$ has to be made greater than that of some message $m'$, where $m'.ts = (rtc', seq')$. The timestamp of $m$ would be changed to $(rtc', seq'+1)$.

In reply to the barrier request for some message $m : m.ts = (rtc,seq)$, a process in $blockers(m)$ would enqueue some message $b : b.ts = (rtc + 1, \text{---})$ to be sent to $m.src$. The whole purpose of this is: if two messages $m$ and $m'$, with initial timestamps $m.ts < m'.ts$ had barriers requested at the same time as their creation, even if $m.ts$ is changed because of $m'$, the barrier received for the delivery of $m'$ will probably also serve $m$ -- except when \mbox{$blockers(m) \setminus blockers(m') \neq \emptyset$} or when an empty message is enqueued, since no barrier is requested for empty messages, but they can still cause timestamp changes in other messages

@@

Some messages may need to have their initial timestamp changed as a possible consequence of the optimistic assumption \emph{A2} not holding. Thus it may be necessary to request new, higher barrier values, resulting in longer delivery latencies, not only for the messages whose timestamp changed, but also for those after them in the delivery queue. Besides, delivering messages out of the initial timestamp order may increase the number of mistakes made by the optimistic delivery. A simple way to eliminate the possibility of having to change timestamps is by discarding messages that arrived too late, that is, with a timestamp lower than that of some already delivered message.

Although discarding messages which do not follow the initial timestamp order conflicts with the properties we have defined for the \amcast{} primitive, we describe this approach here because it brings the benefit of decreased number of mistakes and lower message delivery latency. The validity property would have to be changed to:

\begin{itemize}
  \item[(i)] If a correct process $p$ \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then  every correct process $q \in m.dst$ \amdel{}s $m$ \emph{(optimistic uniform validity)}.
\end{itemize}

%\textbf{Optimistic Uniform Validity}: If a process \amcast{}s $m$ and the optimistic assumption \emph{A2} holds, then one of the correct processes that is a destination of $m$ eventually \consdel{}s $m$.
